Practical Considerations for Lomb-Scargle Periodic Analysis

### 2016-11-15 paper.tex
update continuous discussion

\newcommand{\escience}{1}
\newcommand{\uwastro}{2}
\author{Jacob T. VanderPlas\altaffilmark{\escience}}
\altaffiltext{\escience}{eScience Institute, University of Washington}

@@

The Lomb-Scargle periodogram is a classic algorithm for detecting periodicity
in unevenly-sampled time-series, particularly within the astronomy community.
For example, consider the data showing in \Fig{LINEAR_data}.
This is an irregularly-sampled timeseries showing a single object from the
LINEAR survey, with magnitude measured 280 times over the course of five and
a half years.

It is clear that this object varies with an amplitude of around 0.6 magnitudes.
Computing the Lomb-Scargle periodogram for the data gives us a measure of the
power as a function of period of oscillation (\fig{LINEAR_power}, left), from
which we can determine the period of oscillation of approximately 2.58 hours.
The right panel of \fig{LINEAR_power} shows a folded visualization of
the same data as \fig{LINEAR_data} -- i.e.{} plotted as a function of phase
rather than time.

Often this is exactly how the Lomb-Scargle periodogram is presented: as a clean
procedure to detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be considered
when applying a Lomb-Scargle analysis to real-world datasets.
This paper contains a practical guide to the effective use of the Lomb-Scargle
periodogram for analyzing real-world time-series.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{fig02_LINEAR_PSD}
\caption{{\it Left panel:} the Lomb-Scargle periodogram computed from the data
    in \fig{LINEAR_data}, with an inset detailing the interesting region.
    {\it Right panel:} the input data in \fig{LINEAR_data}, folded over the
    2.58-hour period to show the coherent periodic variability.
    \figlabel{LINEAR_power}
}
\end{figure}

\section{Background: Periods in Continuous Signals}

for a continuously defined signal $g(t)$, the Fourier transform is $\hat{g}(\nu)$,

\begin{equation}
    \mathcal{F}\left\{g\right\}(f) = \int_{-\infty}^\infty g(t) e^{-2\pi i f t} dt
\end{equation}

[Perhaps put this info in an appendix?]

Continuous signal $g(t)$:

\begin{equation}
  \mathcal{F}\left\{g\right\}(f) = \int_{-\infty}^\infty g(t) e^{-2\pi i f t} dt
\end{equation}

The inverse of this operation is

\begin{equation}
  \mathcal{F}^{-1}\left\{g\right\}(f) = \int_{-\infty}^\infty g(t) e^{+2\pi i f t} dt
\end{equation}

We can denote this

\begin{equation}
  \hat{g} = \mathcal{F}\left\{g\right\}
\end{equation}

\subsection{Important Properties of the CFT}

\subsubsection{Linearity}
The Fourier Transform is a linear operation, meaning that for two functions $f(t)$ and $g(t)$ and a scalar value $A$,

\begin{eqnarray}
  \mathcal{F}\{g + h\} &=& \mathcal{F}\{g\} + \mathcal{F}\{h\}\\
  \mathcal{F}\{A g\} &=& A\mathcal{F}\{g\}
\end{eqnarray}

These properties follow straightforwardly from the linearity of the Fourier transform definition.

@@

\begin{equation}
  \{g \ast h\}(t) \equiv \int_{-\infty}^\infty g(t - \tau) h(\tau) d\tau
\end{equation}

@@

\begin{equation}
  \mathcal{F}\{g \ast h\} = \mathcal{F}\{g\}\mathcal{F}\{h\}
\end{equation}

>>>>

\newcommand{\escience}{*}
\newcommand{\uwastro}{+}
\author{Jacob T. VanderPlas\altaffilmark{\escience}}
\altaffiltext{\escience}{eScience Institute, University of Washington}

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{fig02_LINEAR_PSD}
\caption{{\it Left panel:} the Lomb-Scargle periodogram computed from the data
    shown in \fig{LINEAR_data}, with an inset detailing the interesting region.
    {\it Right panel:} the input data in \fig{LINEAR_data}, folded over the
    2.58-hour period to show the coherent periodic variability.
    \figlabel{LINEAR_power}
}
\end{figure}

The Lomb-Scargle periodogram \citep{Lomb76, Scargle82}
is a well-known algorithm for detecting periodicity
in unevenly-sampled time-series, particularly within the astronomy community.
For example, consider the data showing in \Fig{LINEAR_data}.
This is an irregularly-sampled timeseries showing a single object from the
LINEAR survey, with magnitude measured 280 times over the course of five and
a half years.

It is clear that this object varies with an amplitude of around 0.6 magnitudes.
Computing the Lomb-Scargle periodogram for the data gives us a measure of the
power as a function of period of oscillation (\fig{LINEAR_power}, left)\footnote{
This periodogram was computed using tools in the AstroPy project \citep{Astropy2013};
code is available at \url{http://github.com/jakevdp/PracticalLombScargle/}
}, from
which we can determine the period of oscillation of approximately 2.58 hours.
The right panel of \fig{LINEAR_power} shows a folded visualization of
the same data as \fig{LINEAR_data} -- i.e.{} plotted as a function of phase
rather than time.

Often this is exactly how the Lomb-Scargle periodogram is presented: as a clean,
well-defined procedure to detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be considered
when applying a Lomb-Scargle analysis to real-world datasets, and I have found
that these are rarely presented together at an introductory level.
This paper seeks to fill that gap, and provide a practical, semi-technical
guide to the effective use of the Lomb-Scargle method of periodic analysis.

\section{Background: Continuous Fourier Transform}

To understand what the Lomb-Scargle periodogram is doing, we will first step
back and review the subject of Fourier analysis of continuous signals.
If we have a continuously-defined signal $g(t)$, we can compute its Fourier
transform $\hat{g}(f)$ using the following integral:
\begin{equation}
    \hat{g}(f) \equiv \int_{-\infty}^\infty g(t) e^{-2\pi i f t} dt
    \eqlabel{FT-def}
\end{equation}
The inverse relationship is given by:
\begin{equation}
    g(t) \equiv \int_{-\infty}^\infty \hat{g}(f) e^{+2\pi i f t} df
    \eqlabel{IFT-def}
\end{equation}
This can be straightforwardly confirmed using the linearity of integration and
the integral definition of the Dirac delta function.
For convenience below, we will define the Fourier transform operator $\mathcal{F}$,
such that
\begin{eqnarray}
    \mathcal{F}\{g\} &=& \hat{g} \\
    \mathcal{F}^{-1}\{\hat{g}\} &=& g
\end{eqnarray}

\subsection{Motivating the Power Spectral Density}

If you explore the properties of the Fourier transform, you will discover
a couple important properties:

\begin{description}
\item[The Fourier Transform of a Sine wave is a delta function at its frequency.]
    That is, if $g(t) = \sin(2\pi f_0 t)$, then
    \begin{equation}
        \hat{g}(f) \propto \delta(f - f_0) - \delta(f + f_0)
    \end{equation}
\item[The Fourier Transform is a Linear Operation.]
    This means that for functions $a$ and $b$ and scalar $A$,
    \begin{eqnarray}
        \mathcal{F}(a + b) &=& \mathcal{F}(a) + \mathcal{F}(b) \\
        \mathcal{F}(A \cdot a) &=& A\cdot\mathcal{F}(a).
    \end{eqnarray}
    These properties follow straightforwardly from the linearity of integration.
\item[The Fourier Transform of a time-shifted function is unchanged up to a phase.]
    That is, if $g_\tau(t) = g(t - \tau)$, then
    \begin{equation}
        \hat{g}_\tau(f) = e^{-2\pi i f \tau}\hat{g}(f)
    \end{equation}
    This can be confirmed via a change of variables in \eq{FT-def}.
\end{description}

Taken together, what this tells us is that if we define the spectral density
\begin{equation}
    \mathcal{P}(g) \equiv \left|\mathcal{F}(g)\right|^2
\end{equation}
we have a quantity that is {\it insensitive to time-shifts} and consists of a
{\it delta function for each sinusoidal frequency within the data}. In other
words, the spectral density {\it P} measures ...

Because of the first two properties, we see that the Fourier transform can essentially
extract sinusoidal components within a signal:

\begin{description}

\item[Convolution Theorem.]
    The convolution theorem states that the
    Fourier transform of a convolution is a product of the Fourier transforms:
    \begin{equation}
      \mathcal{F}\{g \ast h\} = \mathcal{F}\{g\}\mathcal{F}\{h\}
    \end{equation}
    where the convlution operator $\ast$ is defined by
    \begin{equation}
      \{g \ast h\}(t) \equiv \int_{-\infty}^\infty g(t - \tau) h(\tau) d\tau.
    \end{equation}
    This can be confirmed via the linearity of integration and the integral
    definition of the Dirac delta function.

\item[The Fourier Transform of a Gaussian is a Gaussian.]
    The Gaussian function is defined as
    \begin{equation}
    G(t;\sigma) \propto \exp\left[\frac{-t^2}{2\sigma^2}\right].
    \end{equation}
    The Fourier transform of a Gaussian is another Gaussian:
    \begin{equation}
    \mathcal{F}\{G(t;\sigma)\} \propto G(f; \sigma_f)
    \end{equation}
    where $\sigma_f = (2\pi\sigma)^{-1}$. This can be shown by completing the
    square within the integrand.
\end{description}

@@

@@
